Romans 1 by 1. Documenting a population database for the Roman world

### 2015-12-17 source/3/EAGLE_2016_submission_3.tex

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage{supertabular}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

@@

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}

%%%%%%%%
%TITLE:%
%%%%%%%%

@@

\author[VARGA]{Rada Varga}

\address{Babeș-Bolyai University, Cluj-Napoca}

@@

\begin{document}

@@

The database in itself would have three components (see below Fig.~\ref{fig:schemadatabase}), following the best practices in the field
(\citet{doi:10.3200/HMTS.37.1.34-38}): the sources database (with ``facsimile'' transcription of the sources’ text), the central
database (the complete, corrected, integrated, standardized and coded form of the sources database) and the data
releases (destined for on-line usage, allowing easy extraction of data in view of analyses).

@@

\noindent Within this theoretical frame, whose amplitude implies a gradual and long-term approach, we have chosen to start by
building the epigraphical division of the database. Thus, we began developing a project on the middle classes of the
Low Danube provinces and their epigraphic attestations. The database created in this context together with the platform
www.romans1by1.com represent the skeleton of the future population database of the Roman Empire.

@@

\noindent The core of the database’s epigraphical division is formed by a table used for recording personal
individual data (labelled Personal data), around which the entire network of components needed to ensure proper
information recording is built. However, when starting data entry, one must begin by recording information about the
source in use.

The first table to be filled in is the file of the source – Inscriptions. First of all, each inscription gets a source
code, formed of 5 digits and a symbol/acronym of the province’s name (D for Dacia, MS for Moesia Superior, DAL for
Dalmatia, etc.) – so we have, for example 00001MS. The Inscriptions table has text fields, as well as value lists.
Then, we have text strings for: Relevant expressions, Stylistic details, Atypical features, Observations, Place of
discovery, Place of provenience, Ancient name provenience, Timestamp/Timeframe and External links. Although we are
aware that some of the data (Timestamp/Timeframe) could have benefitted from a standardized form, at this point we
opted for vaster possibilities of expression and adaptation. Other information will be filled using value lists, as
standardization is more suited: Type of inscription, Language, Material. For these we will use the Eagle
vocabularies,\footnote{\url{http://www.eagle-network.eu/resources/vocabularies}} because they offer an already
standardized language.

The Coordinates table links each inscription with the latitude and longitude of its place of provenience, in order to
place it on a map.

The Inscription bibliography was conceived so that extracting complete or selective bibliographical lists would be
possible. Thus, a normalization table includes all bibliographical titles referred to and from it; through a value
list, one can select the Bibliography abbreviation. The exact reference is presented and detailed in the Details and
Comments text strings. Of course, all data are linked to the Inscription code, selected as well from a value list.

Only after properly documenting the source one can access the main and most complex table, Personal data, where
basically the individual file of each person is created. At this point, each new entry represents a singular epigraphic
attestation of an individual, and a unique ID is generated, which will help link the respective character throughout
the various components of the database and with other database entries. The person is also manually linked to the
source using a value list of the inscriptions’ codes. In the event of one person being attested in
multiple epigraphs, each attestation will represent a new entry, and it will be assigned a new unique ID, which will be
doubled, during linkage procedures, by a common ID for all instances of the same person.

The Personal data table was built to host a large variety of information offered by epigraphs, and its structure has
proved, up to this day, rather stable. However, if need be, it can always be extended in order to accommodate any new
kind of information. A first set of its variables are name-connected: Praenomen, Nomen, Cognomen/Personal name,
Father/Master name, Agnomen, Signum, where each also has a drop-down associated for Ethnicity and the Agnomen and
Signum an Observations string. While we believe the possibilities of actually identifying signa for members of the
middle classes, during the Principate period, is rather reduced, we opted for facilitating their correct registration,
in case they are discovered. Other data regard Natione, Ethnicity, Origo and Domus – if and how they are mentioned in
the inscription. As acknowledged above, some information will be recorded, even if they are deductive and not literally
written in the source: Gender and Juridical status (though the servile one often is literally recorded). For the
latter, we have opted for a check box, which, if checked, opens all the available possibilities. The rest of the fields
accommodate supplementary information, when and how there is the case: Occupation, Collegium, Deities, Age (at death),
Details of life/death and Observations.

The Occupation field requires some special attention, as it has the Occupation code associated with it; as we are trying
to propose a codification of Roman occupations/professions, based on and adapting the HISCO classification
model\footnote{\url{http://historyofwork.iisg.nl/} }. While a raw classification and codification based on HISCO might only
be a slight challenge, finding a theoretical model might prove to be a more elaborated task. What we aim at is
constructing the “metadata” for the codification of Roman professions attested on stone and analyse how deep the
classification can realistically go. Much alike other normalization fields present in the sources component of the
database, Occupation code was implemented because it helps dealing faster with small and medium size samples (up to
hundreds of people) in view of publishing preliminary results which are vital for dissemination and further financial
support of the project.

Two important variables, for statistics and working with the data, are Dedicated by and Dedicated for, which state the
relation of the recorded individual with the epigraph and with other persons mentioned by it. Both appear in the simple
form of check boxes.

Another particular check box, in need of supplementary explanations, is Later. The database was conceived for attested
civilians from the middle classes, but sometimes they are associated in inscriptions with militaries or representatives
of the elites. In this case, we have to register a minimum of data on the later as well, in order to build a wider
image of the characters that we focus on. When checking Later, it opens Status, which at its turn opens the following
options in the form of check boxes: Senator, Knight, Local magistrate, Decurion, Imperial priest, Imperial slave,
Imperial freedman, Military personnel. If checked, each of these boxes opens a Details text box; additionally, Local
magistrate and Decurion open a City/Town value list, while Military personnel, opens Rank and Unit value lists. While
elite members and military personnel will at some point be added to the database, for the currently running project
purposes, their social and professional status, together with a minimum of relevant details and relation with the
recorded individuals represent enough information.

\begin{center}
\tablefirsthead{}
\tablehead{}
\tabletail{}
\tablelasttail{}
\begin{supertabular}{|m{3.711cm}|m{2.941cm}|m{4.046cm}|}
\hline
Field label &
Field id &
Data type\\\hline
Praenomen &
rperson\_praenomen &
Text\\\hline
Ethnicity praenomen &
rperson\_eth\_praen &
Value list\\\hline
Nomen &
rperson\_nomen &
Text\\\hline
Ethnicity nomen &
rperson\_eth\_nom &
Value list\\\hline
Cognomen/Personal name &
rperson\_cognomen &
Text\\\hline
Ethnicity cognomen/personal name &
rperson\_eth\_cogn &
Value list\\\hline
Father/Master name &
rperson\_father &
Text\\\hline
Ethnicity father/master name &
rperson\_eth\_fth &
Value list\\\hline
Agnomen &
rperson\_agnomen &
Text\\\hline
Ethnicity agnomen &
rperson\_eth\_agn &
Value list\\\hline
Observations agnomen &
rperson\_remark\_agn &
Text\\\hline
Signum &
rperson\_signum &
Text\\\hline
Ethnicity signum &
rperson\_eth\_sign &
Value list\\\hline
Observations signum &
rperson\_remark\_sign &
Text\\\hline
Natione &
rperson\_nation &
Text\\\hline
Ethnicity &
rperson\_nat\_ethnicity &
Text\\\hline
Gender &
rperson\_rgender\_id &
Value list\\\hline
Juridical status &
rperson\_jstatus &
Checkbox. \ Opens checkboxes:

Citizen/Libertus/Peregrine/Slave/

Veteranus (opens Unit/Rank - valuelists)\\\hline
Tribus &
rperson\_rtribus\_id &
Text\\\hline
Origo &
rperson\_origo &
Text\\\hline
Domus &
rperson\_domus &
Text\\\hline
Collegium &
rperson\_collegium &
Text\\\hline
Occupation &
rperson\_occupation &
Text\\\hline
Occupation code &
rperson\_occ\_code &
Text\\\hline
Deities &
rperson\_deity &
Text\\\hline
Age &
rperson\_age &
Text\\\hline
Details of life/death &
rperson\_death\_details &
Text\\\hline
Dedicated for &
rperson\_dedicated\_for &
Checkbox\\\hline
Dedicated by &
rperson\_dedicated\_by &
Checkbox\\\hline
Inscription code &
rperson\_rinscription\_ids &
Value list\\\hline
Later &
rperson\_now\_stat &
Checkbox. Opens checkbox: Status\\\hline
Status &
rperson\_status &
Checkbox. Opens checkboxes:

Senator (opens Details – text)/

Knight (opens Details – text)/

Local Magistrate (opens Details – text, City – value list)/

Decurion (opens Details – text, City – value list)/

Imperial priest (opens Details – text)/

Imperial slave (opens Details – text)/

Imperial freedman (opens Details – text)/

Military personnel (opens Details – text, Unit/Rank – value lists)\\\hline
Observations &
rperson\_observation &
Text\\\hline
%\caption{Metadata of the Personal data table}
\end{supertabular}
\label{tab:personaldata}
\end{center}

Based on the personal ID given to each individual, the Relationship table will solely name the relationship between
individuals (A to B and B to A), choosing from a drop-down menu. The relationships have been encoded from the start, in
order to make processing quicker; thus first-degree relationships have 10-codes (101-Husband, 102-Wife), second and
third degree relations 20- and respectively 30-codes, non-family relations were given 40- codes and 50-s for the
unspecified/unreadable relations. Male relations were given odd numbers and female ones – even numbers.

@@

\noindent Romans 1 by 1 is a first step towards a comprehensive and exhaustive electronic resource for the attested population of
the Roman Empire. The following normal steps for expanding and enriching the database is elaborating the fitted
metadata for provincial elites and military personnel epigraphically attested.

>>>>

% IF YOU USE PDFLATEX
\usepackage[utf8x]{inputenc}
% if you write in english and in greek
\usepackage{ucs}
\usepackage[greek,english]{babel}
\languageattribute{greek}{polutoniko}

@@

% dummy text (remove in a normal thesis)
% remove if not necessary
\usepackage{siunitx}
%Natbib for bibliography management
\usepackage[authoryear]{natbib}
% custom commands
\newcommand{\bs}{\textbackslash}
\usepackage{xtab}
%%%%%%%%
%TITLE:%
%%%%%%%%

@@

\author[CLUJ]{Rada Varga\corref{first}}

\address[CLUJ]{Babeș-Bolyai University, Cluj-Napoca}

@@

%\usepackage{longtable}
\begin{document}

@@

The database in itself would have three components (see below Fig.~\ref{fig:schemadatabase}), following the best practices in the field
(\citet{doi:10.3200/HMTS.37.1.34-38}): the \emph{sources database} (with ``facsimile'' transcription of the sources’ text), the \emph{central database} (the complete, corrected, integrated, standardized and coded form of the sources database) and the \emph{data releases} (destined for on-line usage, allowing easy extraction of data in view of analyses).

@@

\noindent Within this theoretical frame, whose amplitude implies a gradual and long-term approach, we have chosen to start by
building the epigraphical division of the database. Thus, we began developing a project on the middle classes of the
Low Danube provinces and their epigraphic attestations. The database created in this context together with the platform
\url{www.romans1by1.com} represent the skeleton of the future population database of the Roman Empire.

@@

\noindent The core of the database’s epigraphical division is formed by a table used for recording personal
individual data (labelled \emph{Personal data}), around which the entire network of components needed to ensure proper
information recording is built. However, when starting data entry, one must begin by recording information about the
source in use.

The first table to be filled in is the file of the source – \emph{Inscriptions}. First of all, each inscription gets a source
code, formed of 5 digits and a symbol/acronym of the province’s name (D for Dacia, MS for Moesia Superior, DAL for
Dalmatia, etc.) – so we have, for example 00001MS. The \emph{Inscriptions} table has text fields, as well as value lists.
Then, we have text strings for: \emph{Relevant expressions}, \emph{Stylistic details}, \emph{Atypical features}, \emph{Observations}, \emph{Place of discovery}, \emph{Place of provenience}, \emph{Ancient name provenience}, \emph{Timestamp/Timeframe} and \emph{External links}. Although we are
aware that some of the data (\emph{Timestamp/Timeframe}) could have benefitted from a standardized form, at this point we
opted for vaster possibilities of expression and adaptation. Other information will be filled using value lists, as
standardization is more suited: \emph{Type of inscription}, \emph{Language}, \emph{Material}. For these we will use the Eagle
vocabularies,\footnote{\url{http://www.eagle-network.eu/resources/vocabularies}} because they offer an already
standardized language.

The \emph{Coordinates} table links each inscription with the latitude and longitude of its place of provenience, in order to
place it on a map.

The \emph{Inscription bibliography} was conceived so that extracting complete or selective bibliographical lists would be
possible. Thus, a normalization table includes all bibliographical titles referred to and from it; through a value
list, one can select the \emph{Bibliography abbreviation}. The exact reference is presented and detailed in the \emph{Details} and
\emph{Comments} text strings. Of course, all data are linked to the \emph{Inscription code}, selected as well from a value list.

Only after properly documenting the source one can access the main and most complex table, \emph{Personal data}, where
basically the individual file of each person is created. At this point, each new entry represents a singular epigraphic
attestation of an individual, and a unique ID is generated, which will help link the respective character throughout
the various components of the database and with other database entries. The person is also manually linked to the
source using a value list of the inscriptions’ codes. In the event of one person being attested in
multiple epigraphs, each attestation will represent a new entry, and it will be assigned a new unique ID, which will be
doubled, during linkage procedures, by a common ID for all instances of the same person.

The \emph{Personal data} table was built to host a large variety of information offered by epigraphs, and its structure has
proved, up to this day, rather stable. However, if need be, it can always be extended in order to accommodate any new
kind of information. A first set of its variables are name-connected: \emph{Praenomen}, \emph{Nomen}, \emph{Cognomen/Personal name},
\emph{Father/Master} name, \emph{Agnomen}, \emph{Signum}, where each also has a drop-down associated for \emph{Ethnicity} and the \emph{Agnomen} and
\emph{Signum} an \emph{Observations} string. While we believe the possibilities of actually identifying \emph{signa} for members of the
middle classes, during the Principate period, is rather reduced, we opted for facilitating their correct registration,
in case they are discovered. Other data regard \emph{Natione}, \emph{Ethnicity}, \emph{Origo} and \emph{Domus} – if and how they are mentioned in
the inscription. As acknowledged above, some information will be recorded, even if they are deductive and not literally
written in the source: \emph{Gender} and \emph{Juridical status} (though the servile one often is literally recorded). For the
latter, we have opted for a check box, which, if checked, opens all the available possibilities. The rest of the fields
accommodate supplementary information, when and how there is the case: \emph{Occupation}, \emph{Collegium}, \emph{Deities}, \underline{Age} (at death),
\emph{Details of life/death} and \emph{Observations}.

The \emph{Occupation} field requires some special attention, as it has the \emph{Occupation code} associated with it; as we are trying
to propose a codification of Roman occupations/professions, based on and adapting the HISCO classification
model\footnote{\url{http://historyofwork.iisg.nl/} }. While a raw classification and codification based on HISCO might only
be a slight challenge, finding a theoretical model might prove to be a more elaborated task. What we aim at is
constructing the ``metadata'' for the codification of Roman professions attested on stone and analyse how deep the
classification can realistically go. Much alike other normalization fields present in the sources component of the
database, \emph{Occupation code} was implemented because it helps dealing faster with small and medium size samples (up to
hundreds of people) in view of publishing preliminary results which are vital for dissemination and further financial
support of the project.

Two important variables, for statistics and working with the data, are \emph{Dedicated by} and \emph{Dedicated for}, which state the
relation of the recorded individual with the epigraph and with other persons mentioned by it. Both appear in the simple
form of check boxes.

Another particular check box, in need of supplementary explanations, is Later. The database was conceived for attested
civilians from the middle classes, but sometimes they are associated in inscriptions with militaries or representatives
of the elites. In this case, we have to register a minimum of data on the later as well, in order to build a wider
image of the characters that we focus on. When checking \emph{Later}, it opens \emph{Status}, which at its turn opens the following
options in the form of check boxes: \emph{Senator}, \emph{Knight}, \emph{Local magistrate}, \emph{Decurion}, \emph{Imperial priest}, \emph{Imperial slave},
\emph{Imperial freedman}, \emph{Military personnel}. If checked, each of these boxes opens a Details text box; additionally, \emph{Local
magistrate} and \emph{Decurion} open a \emph{City/Town} value list, while \emph{Military personnel}, opens \emph{Rank} and \emph{Unit} value lists. While
elite members and military personnel will at some point be added to the database, for the currently running project
purposes, their social and professional status, together with a minimum of relevant details and relation with the
recorded individuals represent enough information. 
\vspace{20pt}

%\begin{table}[!h]
\tablefirsthead{\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline }
\bottomcaption{Metadata of the Personal data table}

%\tablehead{\multicolumn{3}{c}%
%{{\captionsize\bfseries \tablename\ \thetable{} --
%continued from previous page}} \\
%\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline }

\tablehead{\hline \textbf{Field label} & \textbf{Field id} & \textbf{Data type} \\ \hline}

\tabletail{\hline \multicolumn{3}{|r|}{{\textit{Continued on next page}}} \\ \hline}
\tablelasttail{\hline}

{\small
\addtolength{\tabcolsep}{-0.5mm}
%\begin{supertabular*}{\textwidth}{p{3.5cm}lp{3cm}}
\begin{xtabular}{|p{.30\textwidth}|p{.30\textwidth}|p{.30\textwidth}|}
%\begin{tabular*}{p{.30\textwidth}p{.29\textwidth}p{.31\textwidth}}
%\toprule
%\textbf{Field label} & \textbf{Field id} & \textbf{Data type}\\\midrule
Praenomen &
rperson\_praenomen &
Text\\ \hline
Ethnicity praenomen &
rperson\_eth\_praen &
Value list\\ \hline
Nomen &
rperson\_nomen &
Text\\ \hline
Ethnicity nomen &
rperson\_eth\_nom &
Value list\\ \hline
Cognomen/Personal name &
rperson\_cognomen &
Text\\ \hline
Ethnicity cognomen/personal name &
rperson\_eth\_cogn &
Value list\\ \hline
Father/Master name &
rperson\_father &
Text\\ \hline
Ethnicity father/master name &
rperson\_eth\_fth &
Value list\\ \hline
Agnomen &
rperson\_agnomen &
Text\\ \hline
Ethnicity agnomen &
rperson\_eth\_agn &
Value list\\ \hline
Observations agnomen &
rperson\_remark\_agn &
Text\\ \hline
Signum &
rperson\_signum &
Text\\ \hline
Ethnicity signum &
rperson\_eth\_sign &
Value list\\ \hline
Observations signum &
rperson\_remark\_sign &
Text\\ \hline
Natione &
rperson\_nation &
Text\\ \hline
Ethnicity &
rperson\_nat\_ethnicity &
Text\\ \hline
Gender &
rperson\_rgender\_id &
Value list\\ \hline
Juridical status &
rperson\_jstatus &
Checkbox. \ Opens checkboxes:

Citizen, Libertus, Peregrine, Slave 
Veteranus (opens Unit/Rank - valuelists)\\ \hline
Tribus &
rperson\_rtribus\_id &
Text\\ \hline
Origo &
rperson\_origo &
Text\\ \hline
Domus &
rperson\_domus &
Text\\ \hline
Collegium &
rperson\_collegium &
Text\\ \hline
Occupation &
rperson\_occupation &
Text\\ \hline
Occupation code &
rperson\_occ\_code &
Text\\ \hline
Deities &
rperson\_deity &
Text\\ \hline
Age &
rperson\_age &
Text\\ \hline
Details of life/death &
rperson\_death\_details &
Text\\ \hline
Dedicated for & rperson\_dedicated\_for & Checkbox\\ \hline
Inscription code & rperson\_rinscription\_ids & Value list\\ \hline
Later &
rperson\_now\_stat & Checkbox. Opens checkbox:
Status\\ \hline
Status & rperson\_status & Checkbox. Opens checkboxes: 
Senator (opens Details – text),
Knight (opens Details – text) 
Local Magistrate (opens Details – text, City – value list), 
Decurion (opens Details – text, City – value list), 
Imperial priest (opens Details – text), 
Imperial slave (opens Details – text), 
Imperial freedman (opens Details – text), 
Military personnel (opens Details – text, Unit/Rank – value lists)\\ \hline
Observations &
rperson\_observation &
Text\\ \hline
%\bottomrule

\end{xtabular}}
%\caption{This is a wide table where we have used the \texttt{tabular*} environment. Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.}
%\label{tab:personaldata}
%\end{table}

%\caption{Metadata of the Personal data table}
%\end{supertabular}
%\label{tab:personaldata}
%\end{center}

Based on the personal ID given to each individual, the \emph{Relationship} table will solely name the relationship between
individuals (A to B and B to A), choosing from a drop-down menu. The relationships have been encoded from the start, in
order to make processing quicker; thus first-degree relationships have 10-codes (101-Husband, 102-Wife), second and
third degree relations 20- and respectively 30-codes, non-family relations were given 40- codes and 50-s for the
unspecified/unreadable relations. Male relations were given odd numbers and female ones – even numbers.

@@

\noindent \emph{Romans 1 by 1} is a first step towards a comprehensive and exhaustive electronic resource for the attested population of
the Roman Empire. The following normal steps for expanding and enriching the database is elaborating the fitted
metadata for provincial elites and military personnel epigraphically attested.
